\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-05-25 move_schedules.tex
Updated declarations in prep for APCT

@@

@@

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagrams of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

@@

The work presented here seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

@@

By making stronger assumptions on the underlying topological space, restricting the homological dimension, or targeting  a weaker invariant (e.g. Betti numbers), one can usually obtain faster algorithms.
For example, Attali et al.~\cite{attali2009persistence} give a linear time algorithm for computing persistence on graphs.
In the same paper, they describe how to obtain $\epsilon$-simplifications of $1$-dimensional persistence diagrams for filtered $2$-manifolds by using duality and symmetry theorems. 
Along a similar vein, Edelsbrunner et al.~\cite{edelsbrunner2000topological} give a fast incremental algorithm for computing persistent Betti numbers up to dimension $2$, again by utilizing symmetry, duality, and ``time-reversal''~\cite{delfinado1995incremental}. Chen et al.~\cite{chen2013output} give an output-sensitive method for computing persistent homology,  utilizing the property that certain submatrices of  $D$ have the same rank as  $R$, which they exploit through fast sub-cubic rank algorithms specialized for sparse-matrices.

@@

@@

Cohen-steiner et al.~\cite{cohen2007stability} referred to these curves as \emph{vines}, a collection of which forms as \emph{vineyard}---the geometric analogy meant to act as a guidepost for practitioners seeking to understand the evolution of topological structure over time.

@@

For completeness, we recapitulate the motivation of the moves algorithm from~\cite{busaryev2010tracking}. 
Let $f: K \rightarrow [m]$ denote a filtration of size $m = \lvert K \rvert$ and $R = DV$ its decomposition, where $R = [r_1, r_2, \dots, r_m]$ and $V = [v_1, v_2, \dots, v_m]$ denote the columns of $R$ and $V$, respectively. By definition, if $r_j = D v_j = 0$, then the inclusion $K_{j-1} \hookrightarrow K_{j}$ introduced a new cycle whose generator is given by:
%the non-zero coefficients $(\alpha_i^{(j)}, \alpha_{i+1}^{(j)}, \dots, \alpha_j^{(j)})$ via the combination: 
\begin{equation}\label{eq:coefficients}
	r_j = D v_j = \alpha_i^{(i)} \cdot \partial(\sigma_{i}) + \alpha_{i}^{(i+1)} \cdot \partial(\sigma_{i+1}) + \dots + \alpha_{i}^{(j)} \cdot \partial(\sigma_{j}) = 0
\end{equation}
Now, consider the permutation $P$ that moves a simplex $\sigma_i$ in $K$ to position $j$, shifting all intermediate simplices $\sigma_{i+1}, \dots, \sigma_{j}$ down by one ($i < j$). 
To perform this shift, the non-zero coefficients from~\eqref{eq:coefficients} must be set to zero, otherwise $\partial(\sigma_i)$ contributes to boundary chains $r_{i+1}, \dots, r_j$ earlier in the filtration. 
As these coefficients are recorded by $v_j$, setting $\alpha_i^{(k)} = 0$  amounts to setting the row entries $v_k(i) = 0$ for all $k \in [i+1,j]$---these cancellations manifest as column operations $(\ast)$ used to ensure $P V(\ast)$ is upper-triangular, thus maintaining invariant I2.
Notice that these operations may yield an unreduced $R' = P R(\ast) P^T$, breaking invariant I3.
We could reduce $R'$ with an additional $k$ operations, with the possibility that $k \sim O(\lvert i - j\rvert^2)$, which is no more efficient than re-running the reduction algorithm on columns $[i,j]$ in $R$.

@@

In the following subsections, we investigate how to  leverage the increased flexibility of the move framework, beginning with the hypothesis formed experimentally from section~\ref{sec:motivation} that vineyards exhibits extraneous overhead maintaining the decomposition.
In theory, decreasing the number of times the decomposition is restored to a valid state ought to reduce this overhead, motivating the question: can one simultaneously minimize the number of times the decomposition is restored to a valid state while retaining an efficient update scheme? 
Next we will show how the analysis of the 
continuous time setting leads to the idea of a schedule, 
and then how a schedule comprised of move operations 
can reduce the overhead from the Vineyards algorithm.

\subsection{Continuous setting}\label{sec:continuous_setting}
In the continuous time case, we are given a homotopy $F : K \times [0,1] \to \mathbb{R}$ interpolating between  a pair of filtrations $f,f' : K \rightarrow [m]$. That is,  the function $t\mapsto F(\sigma, t)$ is continuous,  $f(\sigma) = F(\sigma, 0)$ and $f'(\sigma) = F(\sigma, 1)$ for every  $\sigma \in K$. 
As the choice of homotopy $F$ determines the number of distinct filtrations while deforming $f$ into $f'$ via $F$, bounding this number requires assumptions on $F$. 
Indeed, if we assume that each pair of curves $\big(t,  F(\cdot, t)\big) \subset [0,1]\times \mathbb{R}$  intersect in at most one point, at which they cross, then the  functions $t\mapsto F(\cdot, t)$ belong to a class of $x$-monotone curves called \emph{pseudo-segments}.
%, i.e. parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant. 
This family includes the straight-line homotopy $F(\sigma, t) = (1 - t) f(\sigma) + t f'(\sigma)$,  studied in the original vineyards paper~\cite{cohen2006vines}. 
If there are no crossings in the interval $(a,b)$,
then the filtration order given by $\sigma \mapsto F(\sigma, t)$
is the same for every $t\in (a,b)$.
Thus, changes to the filtration occur only after a crossing event. 
Detecting all $k$ intersections of $m$ pseudo-segments is a well-studied problem in computational geometry that can be optimally solved in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term.

Let $F : K \times [0,1] \to \mathbb{R}$ be a homotopy of pseudo-segments between the  filtrations 
$f,f' : K \rightarrow [m]$.
The continuity and genericity assumptions on $F$ imply that for $ \sigma,\mu \in K$ distinct, 
 the curves   $t \mapsto F(\sigma ,t)$ and 
$t \mapsto F(\mu,t)$ intersect if and only if
 $f(\sigma) > f(\mu)$ and $f'(\sigma)  < f'(\mu)$, or $f(\sigma) < f(\mu)$ and $f'(\sigma)  > f'(\mu)$. 
 In other words, the number of crossings in $F$ is 
 exactly the  \emph{Kendall}-$\tau$ distance~\cite{diaconis1977spearman} between $f$ and $f'$: 
\begin{equation}\label{eq:kendall_dist}
	K_\tau(f, f') = 
 \frac{1}{2} \left\lvert \, \left\{(\sigma, \mu) \mid \mathsf{sign}\big(f(\sigma) - f(\mu) \big) \neq \mathsf{sign}\big(f'(\sigma) - f'(\mu)\big) \right\} \, \right\rvert
\end{equation}
After slightly perturbing $F$ if necessary,  we can further assume that its crossings occur at  $k = K_\tau(f,f')$ distinct time points $0< t_1 < \cdots < t_k < 1$. 
Let $t_0 = 0$, $t_{k+1}= 1$ and fix 
$a_i \in (t_{i}, t_{i + 1})$ for $i=0, \ldots, k$. 
Then, the order in $K$ induced by $\sigma \mapsto F(\sigma, a_i)$
defines a filtration $f_i : K \rightarrow [m]$
so  that $f_0 = f$, $f_k = f'$ and 
$\mathcal{F} = (f_0, f_1,\ldots, f_k)$ is the ordered sequence of all distinct filtrations in the interpolation from $f$ to $f'$ via $F$.

The continuity of the curves $t\mapsto F(\cdot, t)$ and the fact that $t_i$ is the sole crossing time  in the interval $(t_{i-1}, t_{i+1})$, imply that the
permutation $\rho_i $   
transforming $f_{i-1}$ into $f_i$, i.e. so that  $f_i = \rho_i \circ f_{i-1}$, must be (in cycle notation) of the form 
$\rho_i = (\ell_i \;\; \ell_i +1 )$ for $1 \leq \ell_i < m$. In other words, $\rho_i$ is an adjacent transposition for each $i =1,\ldots, k$.
The ordered sequence of adjacent transpositions  
$S_F = (\rho_1, \rho_2, \ldots, \rho_k)$ can thus be thought of as a \emph{schedule} of permutations to be applied to the initial decomposition $R = DV$ of $(K,f)$, in order to   interpolate from $f$ to $f'$ via $F$. 
Going beyond transpositions, the idea of a schedule can be generalized as follows:
\begin{definition}[Schedule]\label{def:schedule}
Given a pair of filtrations $(K, f), (K,f')$ 
%and a fixed bijection $\rho : f \to f'$, let $\rho^\ast : [m] \to [m]$ denote its induced bijection on $S_m$ 
and $R = DV$ the initial decomposition of $(K, f)$, a \emph{schedule}  $\mathcal{S} = (s_1, s_2, \dots, s_d )$ is a sequence of permutations  satisfying: 
	\begin{equation}\label{eq:rv_seq_schedule}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where, for each $i \in [d]$, $R_i = D_i V_i$ is a valid decomposition respecting invariants~\ref{inv:decomposition}, and $R'$ is a valid decomposition for $(K,f')$.
\end{definition}

The size of the schedule $S_F$ defined from the homotopy $F$ above 
is exactly $K_\tau(f,f')$. In the worst case this
is $\sim O(m^2)$, achieved when $f' = - f$, which the grayscale image data example from section~\ref{sec:motivation} exhibits. 
Indeed,  the Freudenthal triangulation of the $9\times 9$ grid contains $(81, 208, 128)$ simplices of dimensions $(0, 1, 2)$, respectively. 
Therefore, $m = 417$ and $\lvert S_F \rvert \leq \frac{1}{2}m(m-1) = 86,736$.
As the homotopy given by the video is simulated,  $\approx 70,\!000$ transpositions are generated, approaching the worst case upper bound due to the fact that $f'$  is nearly the reverse of $f$.

If our goal is to decrease  $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by collapsing contiguous sequences of adjacent transpositions to moves, via the map:
\begin{equation}\label{eq:tr_to_mv}
	 (i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)  \quad \text{ if } i < j
\end{equation}
Clearly $\lvert \widetilde{S}_F \rvert \leq \lvert S_F \rvert$ and the associated coarsened $\widetilde{S}_F$ requires just $O(m)$ time to compute. However, the coarsening depends entirely on the initial choice of $F$ and the quadratic upper bound remains---it is always possible that there are no contiguous subsequences to collapse. 
As mentioned in~\ref{sec:vineyards}, this quadratic scaling induces a number of issues in the practical implementations of the vineyards algorithm.
% Indeed, although many transpositions require no column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time per transposition.

@@

Contrasting the continuous-time setting, if we discard the use of a homotopy interpolation and allow move operations in any order, we obtain a trivial upper bound of $O(m)$ on the schedule size:  simply move each simplex in $K$ from  its position in the filtration given $f$ to the position given by $f'$---which we call the \emph{naive strategy}. However, it is not immediately clear whether this bound is tight. In this section, we investigate the tightness of this bound by revisiting the problem from a combinatorial perspective. 
% or whether this strategy of generating schedules---which we call \emph{naive strategy}---is performant.  
%Indeed, arbitrarily permuting a given filtration $K$: one must respect the face poset.

@@

Indeed, Diaconis et al.~\cite{diaconis1977spearman} showed the Spearman distance approximates $K_\tau$ within a factor of two: 

@@

\subsubsection{Fibered barcode}\label{sec:fibered_barcode}
Unfortunately, unlike the one-parameter case, there is no complete discrete invariant for multi-parameter persistence.
Circumventing this, Lesnick et al~\cite{lesnick2015interactive} associate a variety of incomplete invariants to 2-parameter persistence modules; we focus here on the \emph{fibered barcode} invariant, defined as follows: 
\begin{definition}[Fibered barcode]
	%\in \overline{\mathcal{L}}
	The fibered barcode $\mathcal{B}(M)$ of a 2D persistence module $M$ is the map which sends each line  $L \subset \mathbb{R}^2$ with non-negative slope to the barcode $\mathcal{B}_L(M)$: 
%	$$ L \mapsto \mathcal{B}_L(M) $$
$$ \mathcal{B}(M) = \{ \; B_L(M) : L \in \mathbb{R} \times \mathbb{R}^{+} \; \}$$
Equivalently, $\mathcal{B}(M)$ is the 2-parameter family of barcodes given by restricting $M$ to the of set affine lines with non-negative slope in $\mathbb{R}^2$. 
\end{definition} 
\noindent Although an intuitive invariant, it is not clear how one might go about computing $\mathcal{B}(M)$ efficiently. 
One obvious choice is fix $L$ via a linear combination of two filter functions, restrict $M$ to $L$, and compute the associated 1-parameter barcode. 
However, this is an $O(m^3)$ time computation, which is prohibitive for exploratory data analysis purposes.

@@

After Lee et al published their work, Carlsson et al.~\cite{carlsson2008local} subsequently performed extensive experiments using persistent homology, wherein he found that the distribution of high-contrast $3 \times 3$ patches is actually well-approximated by a Klein bottle $\mathcal{M}$---around 60\% of the high-contrast patches from the still image data set lie within a small neighborhood around $\mathcal{M}$ which accounts for only 21\% of the 7-spheres volume. Along a similar vein in the sparse coding context, Perea et al.~\cite{perea2014klein} introduced a dictionary learning framework for estimating the distribution of patches from texture images.

@@

>>>>

%\usepackage[nodisplayskipstretch]{setspace}
%\setstretch{1.5}

@@

\usepackage{setspace}
\setstretch{1}
%\setlength{\abovedisplayskip}{0pt}
%\setlength{\belowdisplayskip}{0pt}
%\setlength{\abovedisplayshortskip}{0pt}
%\setlength{\belowdisplayshortskip}{0pt}

@@

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagrams of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, the initial reduction cost is amortized by the cost of maintaining the decomposition, implying each diagram produced by vineyards takes just linear time per time point. 
% approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

@@

The work presented here seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
Though the vineyards algorithm is efficient at constructing a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ For all accompanying software and materials, see: \url{https://github.com/peekxc/move_schedules}}

@@

By making stronger assumptions on the underlying topological space, restricting the homological dimension, or targeting  a weaker invariant (e.g. Betti numbers), one can usually obtain faster algorithms.
For example, Attali et al.~\cite{attali2009persistence} give a linear time algorithm for computing persistence on graphs.
In the same paper, they describe how to obtain $\epsilon$-simplifications of $1$-dimensional persistence diagrams for filtered $2$-manifolds by using duality and symmetry theorems. 
Along a similar vein, Edelsbrunner et al.~\cite{edelsbrunner2000topological} give a fast incremental algorithm for computing persistent Betti numbers up to dimension $2$, again by utilizing symmetry, duality, and ``time-reversal''~\cite{delfinado1995incremental}. Chen \& Kerber~\cite{chen2013output} give an output-sensitive method for computing persistent homology,  utilizing the property that certain submatrices of  $D$ have the same rank as  $R$, which they exploit through fast sub-cubic rank algorithms specialized for sparse-matrices.

@@

where $\mathrm{LCS}(f,f')$ refers to the size of the longest common subsequence between the simplexwise filtrations $(K, f)$ and $(K,f')$ (see section~\ref{sec:moves_setting} for more details).

@@

Cohen-Steiner et al.~\cite{cohen2007stability} referred to these curves as \emph{vines}, a collection of which forms as \emph{vineyard}---the geometric analogy meant to act as a guidepost for practitioners seeking to understand the evolution of topological structure over time.

@@

For completeness, we recapitulate the motivation of the moves algorithm from~\cite{busaryev2010tracking}. 
Let $f: K \rightarrow [m]$ denote a filtration of size $m = \lvert K \rvert$ and $R = DV$ its decomposition, where $R = [r_1, r_2, \dots, r_m]$ and $V = [v_1, v_2, \dots, v_m]$ denote the columns of $R$ and $V$, respectively. By definition, if $r_j = D v_j = 0$, then the inclusion $K_{j-1} \hookrightarrow K_{j}$ introduced a new cycle whose generator is given by:
%the non-zero coefficients $(\alpha_i^{(j)}, \alpha_{i+1}^{(j)}, \dots, \alpha_j^{(j)})$ via the combination: 
\begin{equation}\label{eq:coefficients}
	r_j = D v_j = \alpha_j^{(1)} \cdot \partial(\sigma_{1}) + \alpha_{j}^{(2)} \cdot \partial(\sigma_{2}) + \dots + \alpha_{j}^{(j)} \cdot \partial(\sigma_{j}) = 0
\end{equation}
Now, consider the permutation $P$ that moves a simplex $\sigma_i$ in $K$ to position $j$, shifting all intermediate simplices $\sigma_{i+1}, \dots, \sigma_{j}$ down by one ($i < j$). 
To perform this shift, all of the non-zero coefficients $\alpha_k^{(i)}$ contributing to $Dv_k$ for $k \in [i+1, j]$ must be set to zero, as otherwise $\partial(\sigma_i)$ may participate in boundary chains earlier than its own birth in the permuted filtration. 
Setting these coefficients to zero amounts to setting row entries $v_k(i) = 0$ for all $k \in [i+1,j]$---these cancellations manifest as column operations $(\ast)$ used to ensure $P V(\ast)$ is upper-triangular, thus maintaining invariant I2.
Notice that these operations may yield an unreduced $R' = P R(\ast) P^T$, breaking invariant I3.
We could reduce $R'$ with an additional $k$ operations, with the possibility that $k \sim O(\lvert i - j\rvert^2)$, but then this is no more efficient than re-running the reduction algorithm on columns $[i,j]$ in $R$.

@@

In the following subsections, we investigate how to  leverage the increased flexibility of the move framework, beginning with the hypothesis formed experimentally from section~\ref{sec:motivation} that vineyards exhibits extraneous overhead maintaining the decomposition.
In theory, decreasing the number of times the decomposition is restored to a valid state ought to reduce this overhead, motivating the question: can one simultaneously minimize the number of times the decomposition is restored to a valid state while retaining an efficient update scheme? 
%Next we will show how the analysis of the 
%continuous time setting leads to the idea of a schedule, 
%and then how a schedule comprised of move operations 
%can reduce the overhead from the Vineyards algorithm.

\subsection{Continuous setting}\label{sec:continuous_setting}
In the original vineyards setting, a given homotopy  $F : K \times [0,1] \to \mathbb{R}$ continuously interpolating between $(K, f)$ and $(K, f')$ is discretized into a set of critical events that alter the filtration order. 
As $F$ determines the number of distinct filtrations encountered during the deformation from $f$ to $f'$, a natural question is whether such an interpolation can be modified so as to minimize the number of times the decomposition is restored to a valid state.
Towards explaining the phenomenon exhibited in Figure~\ref{fig:vineyards}, in what follows we analyze a particular class of interpolation schemes in order to establish an upper bound on this quantity.

%a homotopy $F : K \times [0,1] \to \mathbb{R}$  whose function $t\mapsto F(\sigma, t)$ is continuous and satisfies $f(\sigma) = F(\sigma, 0)$ and $f'(\sigma) = F(\sigma, 1)$ for every  $\sigma \in K$. 
%Indeed, if we assume that each pair of curves $\big(t,  F(\cdot, t)\big) \subset [0,1]\times \mathbb{R}$  intersect in at most one point, at which they cross, then the  functions $t\mapsto F(\cdot, t)$ belong to a class of $x$-monotone curves called \emph{pseudo-segments}.
%%, i.e. parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant. 
%This family includes the straight-line homotopy $F(\sigma, t) = (1 - t) f(\sigma) + t f'(\sigma)$,  studied in the original vineyards paper~\cite{cohen2006vines}. 
%If there are no crossings in the interval $(a,b)$,
%then the filtration order given by $\sigma \mapsto F(\sigma, t)$
%is the same for every $t\in (a,b)$.
%Thus, changes to the filtration occur only after a crossing event. 
%Detecting all $k$ intersections of $m$ pseudo-segments is a well-studied problem in computational geometry that can be optimally solved in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term.

%Necessarily, bounding this quantity requires assumptions on $F$. 
Let $F : K \times [0,1] \to \mathbb{R}$ be a homotopy of $x$-monotone curves\footnote{This term has been used in reference to parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant, see~\cite{boissonnat2000efficient}. } between the  filtrations 
$f,f' : K \rightarrow [m]$ whose function $t\mapsto F(\sigma, t)$ is continuous and satisfies $f(\sigma) = F(\sigma, 0)$ and $f'(\sigma) = F(\sigma, 1)$ for every  $\sigma \in K$.
Note this family includes the straight-line homotopy $F(\sigma, t) = (1 - t) f(\sigma) + t f'(\sigma)$,  studied in the original vineyards paper~\cite{cohen2006vines}. 
If we assume that each pair of curves $\big(t,  F(\cdot, t)\big) \subset [0,1]\times \mathbb{R}$  intersect in at most one point---at which they cross---the continuity and genericity assumptions on $F$ imply that for $ \sigma,\mu \in K$ distinct,  the curves   $t \mapsto F(\sigma ,t)$ and $t \mapsto F(\mu,t)$ intersect if and only if $f(\sigma) > f(\mu)$ and $f'(\sigma)  < f'(\mu)$, or $f(\sigma) < f(\mu)$ and $f'(\sigma)  > f'(\mu)$. 
 In other words, the number of crossings in $F$ is 
 exactly the  \emph{Kendall}-$\tau$ distance~\cite{diaconis1977spearman} between $f$ and $f'$: 
\begin{equation}\label{eq:kendall_dist}
	K_\tau(f, f') = 
 \frac{1}{2} \left\lvert \, \left\{(\sigma, \mu) \mid \mathsf{sign}\big(f(\sigma) - f(\mu) \big) \neq \mathsf{sign}\big(f'(\sigma) - f'(\mu)\big) \right\} \, \right\rvert
\end{equation}
After slightly perturbing $F$ if necessary,  we can further assume that its crossings occur at  $k = K_\tau(f,f')$ distinct time points $0< t_1 < \cdots < t_k < 1$. 
Let $t_0 = 0$, $t_{k+1}= 1$ and fix 
$a_i \in (t_{i}, t_{i + 1})$ for $i=0, \ldots, k$. 
Then, the order in $K$ induced by $\sigma \mapsto F(\sigma, a_i)$
defines a filtration $f_i : K \rightarrow [m]$
so  that $f_0 = f$, $f_k = f'$ and 
$\mathcal{F} = (f_0, f_1,\ldots, f_k)$ is the ordered sequence of all distinct filtrations in the interpolation from $f$ to $f'$ via $F$.

The continuity of the curves $t\mapsto F(\cdot, t)$ and the fact that $t_i$ is the sole crossing time  in the interval $(t_{i-1}, t_{i+1})$, imply that the
permutation $\rho_i $   
transforming $f_{i-1}$ into $f_i$, i.e. so that  $f_i = \rho_i \circ f_{i-1}$, must be (in cycle notation) of the form 
$\rho_i = (\ell_i \;\; \ell_i +1 )$ for $1 \leq \ell_i < m$. In other words, $\rho_i$ is an adjacent transposition for each $i =1,\ldots, k$.
The ordered sequence of adjacent transpositions  
$S_F = (\rho_1, \rho_2, \ldots, \rho_k)$ can thus be thought of as a \emph{schedule} of permutations to be applied to the initial decomposition $R = DV$ of $(K,f)$, in order to   interpolate from $f$ to $f'$ via $F$. 
Going beyond transpositions, the idea of a schedule can be generalized as follows:
\begin{definition}[Schedule]\label{def:schedule}
Given a pair of filtrations $(K, f), (K,f')$ 
%and a fixed bijection $\rho : f \to f'$, let $\rho^\ast : [m] \to [m]$ denote its induced bijection on $S_m$ 
and $R = DV$ the initial decomposition of $(K, f)$, a \emph{schedule}  $\mathcal{S} = (s_1, s_2, \dots, s_d )$ is a sequence of permutations  satisfying: 
	\begin{equation}\label{eq:rv_seq_schedule}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where, for each $i \in [d]$, $R_i = D_i V_i$ is a valid decomposition respecting invariants~\ref{inv:decomposition}, and $R'$ is a valid decomposition for $(K,f')$.
\end{definition}
\noindent Observe the size of the schedule $S_F$ defined from the homotopy $F$ above is exactly $K_\tau(f,f')$.
On the positive side, the reduction of schedule planning to crossing detection implies the former can be  that can be solved optimally in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term and $m$ is the number of simplices in the filtration(s).
On  the negative side, $k = K_\tau(f,f')$ scales in size to $\sim O(m^2)$ in the worst case, achieved when $f' = - f$. As mentioned in~\ref{sec:vineyards}, this quadratic scaling induces a number of issues in the practical implementations of the vineyards algorithm

%detecting all $k$ intersections of $m$ such curves is a well-studied problem in computational geometry
%If there are no crossings in the interval $(a,b)$,
%then the filtration order given by $\sigma \mapsto F(\sigma, t)$
%is the same for every $t\in (a,b)$.
%Thus, changes to the filtration occur only after a crossing event. 
\begin{remark}
	The grayscale image data example from section~\ref{sec:motivation} exhibits this quadratic scaling.
Indeed, the Freudenthal triangulation of the $9\times 9$ grid contains $(81, 208, 128)$ simplices of dimensions $(0, 1, 2)$, respectively. 
Therefore, $m = 417$ and $\lvert S_F \rvert \leq \frac{1}{2}m(m-1) = 86,736$.
As the homotopy given by the video is simulated,  $\approx 70,\!000$ transpositions are generated, approaching the worst case upper bound due to the fact that $f'$  is nearly the reverse of $f$.  
\end{remark}

If our goal is to decrease  $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by e.g. collapsing contiguous sequences of adjacent transpositions to moves, via the map:
\begin{equation}\label{eq:tr_to_mv}
	 (i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)  \quad \text{ if } i < j
\end{equation}
Clearly $\lvert \widetilde{S}_F \rvert \leq \lvert S_F \rvert$ and the associated coarsened $\widetilde{S}_F$ requires just $O(m)$ time to compute. However, the coarsening depends entirely on the initial choice of $F$ and the quadratic upper bound remains---it is always possible that there are no contiguous subsequences to collapse. 
Suggesting one must either abandon the continuous  setting or make stronger assumptions on $F$ to have any hope of keeping $\lvert S_F \rvert \sim O(m)$ in size.
% Indeed, although many transpositions require no column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time per transposition.

@@

Contrasting the continuous-time setting, if we discard the use of a homotopy interpolation and allow move operations in any order, we obtain a trivial upper bound of $O(m)$ on the schedule size:  simply move each simplex in $K$ from  its position in the filtration given by $f$ to the position given by $f'$---which we call the \emph{naive strategy}. However, in losing the interpolation interpretation it is no longer clear the $O(m)$ bound is tight, and the ``intermediate'' filtrations need no longer respect the face poset of the underlying complex $K$. In this section, we investigate these issues from a combinatorial perspective. 
% or whether this strategy of generating schedules---which we call \emph{naive strategy}---is performant.  
%Indeed, arbitrarily permuting a given filtration $K$: one must respect the face poset.

@@

Indeed, Diaconis~\cite{diaconis1977spearman} showed the Spearman distance approximates $K_\tau$ within a factor of two: 

@@

\subsubsection{Fibered barcode}\label{sec:fibered_barcode}
Unfortunately, unlike the one-parameter case, there is no complete discrete invariant for multi-parameter persistence.
Circumventing this, Lesnick et al~\cite{lesnick2015interactive} associate a variety of incomplete invariants to 2-parameter persistence modules; we focus here on the \emph{fibered barcode} invariant, defined as follows: 
\begin{definition}[Fibered barcode]
	%\in \overline{\mathcal{L}}
	The fibered barcode $\mathcal{B}(M)$ of a 2D persistence module $M$ is the map which sends each line  $L \subset \mathbb{R}^2$ with non-negative slope to the barcode $\mathcal{B}_L(M)$: 
%	$$ L \mapsto \mathcal{B}_L(M) $$
$$ \mathcal{B}(M) = \{ \; B_L(M) : L \in \mathbb{R} \times \mathbb{R}^{+} \; \}$$
Equivalently, $\mathcal{B}(M)$ is the 2-parameter family of barcodes given by restricting $M$ to the of set affine lines with non-negative slope in $\mathbb{R}^2$. 
\end{definition} 
\noindent Although an intuitive invariant, it is not clear how one might go about computing $\mathcal{B}(M)$ efficiently. 
One obvious choice is fix $L$ via a linear combination of two filter functions, restrict $M$ to $L$, and compute the associated 1-parameter barcode. 
However, this is an $O(m^3)$ time computation, which is prohibitive for interactive data analysis purposes.

@@

After Lee et al published their work, Carlsson et al.~\cite{carlsson2008local} subsequently performed extensive experiments using persistent homology, wherein he found that the distribution of high-contrast $3 \times 3$ patches is actually well-approximated by a Klein bottle $\mathcal{M}$---around 60\% of the high-contrast patches from the still image data set lie within a small neighborhood around $\mathcal{M}$ which accounts for only 21\% of the 7-spheres volume. Along a similar vein in the sparse coding context, Perea~\cite{perea2014klein} introduced a dictionary learning framework for estimating the distribution of patches from texture images.

@@

\section{Declarations}
